Childhood exposure to non-persistent endocrine disruptors, glucocorticosteroids, and neurodevelopment: A study based on the parametric g-formula

### 2024-01-04 output/paper/manuscript.qmd
Finished first draft manuscript w/o revision. Still nothing about sensitivity to period. Added abstract.

>>>>

Results of epidemiological studies seem to indicate that exposure to certain environmental chemicals, including endocrine disruptors (EDCs), is linked to perturbation of the hypothalamic-pituitary-adrenocortical (HPA) axis, which has a major role in brain development. We evaluated the effects of childhood exposure to organophosphate pesticides, phenols, and phthalate metabolites, on urinary glucocorticosteroids and the Attention Network Test (ANT) in childhood, using data from the Human Early Life Exposome (HELIX) Study. We used the parametric g-formula and marginal contrasts to estimate effects between EDCs, glucocorticosteroids, and a measure of efficiency from the ANT, the hit reaction time standard error (HRT-SE), in roughly 1,000 children aged 6-11. We further tested for possible effect modification by sex. We found positive marginal contrasts for exposure to the phenol MEPA and the phthalate metabolites oxo-MiNP, oh-MiNP, and MEHP, on HRT-SE. Multiple EDCs were also associated with positive marginal contrasts of cortisone, cortisol, and corticosterone production. Increased levels of the glucocorticosteroids had no effect on HRT-SE, although we found possible effect modification by sex for cortisone and corticosterone production. Our results suggest that multiple EDCs might interfere with the homeostasis of the HPA axis. Further research is necessary to determine whether this can have a clinically significant impact on childhood neurodevelopment.
